Assessing primitives performance\\ on multi-stage execution

### 2018-05-03 paper.tex
comments

\begin{figure}[tb]
		\centering
		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

@@

%TODO

primitive implemented in C or Cog's back-end.

exec hybrid interp 
 au runtime JIT ST->ASM)
 execution with C and St stack

coût de changement entre la pile St et la pile C et que tu expliques le probleme des primitives C courtes appelées depuis le code généré par le JIT. C'est important pour la suite.

@@

\paragraph{Final specification.} The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.

@@

 \cba{petite intro avec plan de section}

@@

- Pure Smalltalk: baseline
Smalltalk code on top of VM. Used as our baseline for benchmarks.
Write down the Smalltalk code here.
Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:

- Smart Syntax Plugin
%CHECK how to import code env in latex.
%\begin{code}
%= aString 	
%	self == aString ifTrue: [ ^true ].
%	aString isString ifFalse: [ ^false ].
%	self size = aString size ifFalse: [ ^false ].
%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [ ^ false ] ].
%	^ true
%\end{code}
Restricted Smalltalk

written in Smalltalk with restrictions.
Compiled to C AOT into a plugin.

- Slang plugin 
Slang, but no access au fonctions de la VM drectement.
 interpreterProxy. (C compiler inlining, linking time ? cmacro ?)

- Slang pure

- Slang pure + Cog RTL

@@

@@

Baseline performance is not that great, jtted performance is good only with mature optimizing JIT we don't have.

Number of implementations: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
recompilation. 
changing the prim => changing all clients (VM compil from Pharo or Squeak different results
Performance: cross file inlining not possible by default. (InterpreterProxy indirection)
Control on generated code => difference LLVM/GCC
 Pure C performance (no interpreter proxy).
 no C-St jump. Control over the outcome (reliable performance)
 costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers 
 need to recompile vm sources (trampolines), code readability

@@

@@

\subsection{Macro-benchmarks}
JSON

\section{Related work}

>>>>

@@

\begin{figure}[bth]
		\centering
		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

The Cog VM executes the client's code using an hybrid runtime with an interpreter and a JIT. The interpreter uses a global look-up cache to speed-up virtual calls. On cache hits, it requests the JIT to compile that method and the native code version is used instead. In most case, the JIT is used at the second activation of the method. Closures have similar heuristics.

The VM code, including the interpreter code, is compiled by a C compiler and uses the C stack at runtime. During its execution, the interpreter modifies the execution stack of the client language, which is different from its own code stack as shown on Figure \ref{fig:TwoStacks}. This means for example that the frame pointer register in the processor refers to a stack frame on the C stack and that a processor push instruction pushes a value on the C stack. When executing the native code generated by JIT, only the client language stack is used. This means for example that the frame pointer register in the processor refers to a stack frame on the client stack and that a processor push instruction pushes a value on the client stack.

\begin{figure}[tbh]
		\centering
		\includegraphics[width=0.99\linewidth]{figures/TwoStacks}
		\caption{Client and C stacks}
		\label{fig:TwoStacks}
\end{figure}

Because of this design, switching the execution between the interpreter and the native code generated by the JIT has a cost. Effectively, the VM needs to switch the frame pointer, the stack pointer and depending on architecture the link register from one stack to the other one. It also needs to save those pointers to be able to resume execution from one stack to the other one. In addition, caller-saved registers on the C conventions or the JIT convention need to be saved since the VM cannot tell ahead what code is going to be used in the other execution model and what registers are required. We estimate the overhead to up to around a dozen of processor instruction, the exact number varies a lot so it's difficult to precise it (the number of register to edit depends on the processor and on the code executed).

Most of the primitives are implemented either in Slang or directly in C. They benefit from the C compiler optimizations to be efficient. When called from the interpreter, activating a primitive is just a C call (the primitive function pointer is cached in the look-up cache next to the method to activate). When called from the native code generated by the JIT, activating a primitive requires to switch the execution form the client stack to the C stack, perform the primitive, and switch back to resume execution. If the primitive takes a significant amount of instructions to execute, for example, the primitive is copying 100kb from one array to another one, the stack switch dance overhead is negligible. If the primitive is very quick to perform, for example, the primitive is adding two integers which do not overflow, the overhead is massive: the execution cost of the addition goes from a few processor instructions to a few dozen instructions.

\cba{TODO!}
To solve this performance bottleneck, the Cog VM allows to redefine primitives on top of its back-end. 
integrated as part of the native code generated.
No switch cost.
No need to reimpl long running code -> direct in Abs assembly, hard to write and maintain, have to be smart to compete with optimized C code.
=> often partial impl for quick to perform prim, for ex, add of 2 int, if overflow just go back to C prim.

@@

\paragraph{Final specification.} The final specification we productized last month takes two or three parameters. The third parameter, the order, is optional. If the third parameter is absent, the VM assumes the byte ordering (ASCII order) is the order to compare against. Therefore, the VM does not need to go through the indirection order array in the common case. The primitive answers a negative value if the first parameter is smaller than the second, 0 if they are equal, and a positive value if the first parameter is greater than the second one.
\cba{Maybe rewrite the last paragraph to make it clearer / more formal.}

@@

 \cba{petite intro avec plan de section, context was moved before.}

@@

 \cba{Maybe wrte version A, B, C , D , E or a word to represent what they are (Smalltalk) (SmartSyntax) (Plugin) (Slang) (Slang+JIT)
 so you can refer to that to talk about any of these later and not wirte each time : the implem with ths wirtten that wya and exec that way.
 Myabe those could be paragraph titles instead of the following}

\paragraph{Baseline: pure Smalltalk.}
: baseline
Smalltalk code on top of VM. Used as our baseline for benchmarks.
Write down the Smalltalk code here.
Expliquer to:do: over do: (to explain expresiveness) => core lib often use to:do: and not do:

New spec

\paragraph{Existing: Smart Syntax Plugin.}
%CHECK how to import code env in latex.
%Note we're talking about comparison, not =, so not sure we should show = or bench =
%\begin{code}
%= aString 	
%	self == aString ifTrue: [ ^true ].
%	aString isString ifFalse: [ ^false ].
%	self size = aString size ifFalse: [ ^false ].
%	1 to: self size do: [:i | (self at: i) = (aString at: i) ifFalse: [ ^ false ] ].
%	^ true
%\end{code}
Restricted Smalltalk

written in Smalltalk with restrictions.
Compiled to C AOT into a plugin.
Write once, fall-back + compiled.

old spec

\paragraph{Existing: Slang Plugin.}
Slang, but no access au fonctions de la VM drectement.
 interpreterProxy. (C compiler inlining, linking time ? cmacro ?)

 old spec

\paragraph{Pure Slang.}

new spec

\paragraph{Pure Slang with JIT support.}
Cog RTL

@@

new spec

@@

Baseline performance is not that great.

Number of implementations to maintain: 1, 2 or 3: same implem for primitive in C and fall back code in Smalltalk, relatively easy to read/code.
recompilation. 
changing the prim => changing all clients (VM compil from Pharo or Squeak different results
Performance: cross file inlining not possible by default. (InterpreterProxy indirection)
Control on generated code => difference LLVM/GCC
 Pure C performance (no interpreter proxy).
 no C-St jump. Control over the outcome (reliable performance)
 costly jumps between Smalltalk and C runtime (trampolines), need to recompile vm sources, little control over outcome (c compilers 
 need to recompile vm sources (trampolines), code readability

@@

Il faut expliquer en une phrase: 
petit string = cout important C-St stack switch.
Gde string = cout de la boucle.

@@

\subsection{Macro-benchmarks}
Intro sentence with ref to squeak speed center.
JSON

\section{Discussion and related work}

\cba{C'est un peu de la discussion dans le sens que le related work c'est que celui de Tim, le reste c'est plus des solutions pour le pb du stack switch, qui n'est pas le problème résolu dans ce papier}
